Parallel global optimization algorithm employing decision trees for launching local methods

### 2025-09-17 2025/MMST_DecisionTree/ParallelDecisionTree.tex

%Поскольку в реальных задачах глобальной оптимизации каждое вычисление значения функции (далее \textit{trial}) является операцией, требующей больших вычислительных затрат, количество таких попыток приходится сокращать. Это может быть достигнуто за счет преднамеренного выбора вариантов в ходе поиска оптимального решения, отсекая бесперспективные поисковые под области и исследуя только те, в которых можно найти решение задачи. Алгоритм глобального поиска (АГП) основан на этой идее \cite{Strongin2000}. В настоящей работе мы попытались объединить АГП и метод локальной оптимизации \cite{Nocedal, Kelley}  (метод поиска по образцу Хука-Дживса \cite{HookJeeves}), чтобы уменьшить количество выполняемых испытаний. Решение о запуске локального метода будет приниматься с использованием дерева решений.
In real-world global optimization problems, each function evaluation (hereafter referred to as a \textit{trial}) is a computationally expensive operation, making it necessary to reduce the number of such trials. This may be achieved by the purposeful selection of alternatives during the course of optimization, rejecting unpromising search sub-regions and exploring only those where a solution can be found. The Global Search Algorithm (GSA) is based on this idea \cite{Strongin2000}. In this work, we attempted to combine the GSA and a local optimization method \cite{Nocedal, Kelley} (the Hooke-Jeeves pattern search method \cite{HookJeeves}) to reduce the number of trials performed. The decision to initiate the local method will be made using a decision tree.

@@

Classical local optimization algorithms are designed to find a single extremum (local minimum or maximum) in the neighborhood of an initial guess. Depending on the information used about the objective function, these methods are typically classified as:
\begin{itemize}
	\item \textbf{Zero-order methods}, which use only function values;
	\item \textbf{First-order methods} (gradient methods), requiring the computation of first derivatives;
	\item \textbf{Second-order methods} (e.g., Newton's method), which require the computation of the matrix of second derivatives (Hessian).
\end{itemize}

@@

Algorithms terminated when current trial point was within the $\varepsilon$-neighborhood of the true global minimum. Tables \ref{tab:1} and \ref{tab:2} show the average solution time and speedup relative to a sequential run, achieved by the algorithms when solving optimization problems. Figure \ref{fig:fig4} shows the speedup in terms of the number of iterations. Parallelization was implemented using MPI, and experiments were conducted on 1, 8, 16, 32, and 64 processes.

@@

\begin{table}
	\caption{The average solution time}
	\label{tab:1}
	\center
	\begin{tabular}{cccccccccccc}
		\hline\noalign{\smallskip}
		$P$ & \multicolumn{2}{c}{ $N=2$ } & & \multicolumn{2}{c}{$N=3$} & & \multicolumn{2}{c}{$N=4$} & & \multicolumn{2}{c}{$N=5$}  \\
		\noalign{\smallskip} \cline{2-3} \cline{5-6} \cline{8-9} \cline{11-12} \noalign{\smallskip}
		& \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard}  \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
		1   & 6,35 & 12,71 & & 5,92 & 14,19 & & 31,88 & 66,29  & & 53,77 & 324,04 \\
		8   & 0,57 & 1,98  & & 1,63 & 5,59  & & 9,44  & 34,93  & & 6,62 & 52,42 \\
		16  & 1,27 & 1,71  & & 1,17 & 2,10  & & 4,71  & 8,53   & & 3,38 & 17,55 \\
		32  & 0,95 & 0,87  & & 0,38 & 0,75  & & 4,67  & 4,95   & & 1,99 & 11,48 \\
		64  & 0,21 & 1,65  & & 0,41 & 0,60  & & 5,51  & 2,93   & & 1,70 & 10,60 \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

\begin{table}
	\caption{Time speedup}
	\label{tab:2}
	\center
	\begin{tabular}{cccccccccccc}
		\hline\noalign{\smallskip}
		$P$ & \multicolumn{2}{c}{ $N=2$ } & & \multicolumn{2}{c}{$N=3$} & & \multicolumn{2}{c}{$N=4$} & & \multicolumn{2}{c}{$N=5$}  \\
		\noalign{\smallskip} \cline{2-3} \cline{5-6} \cline{8-9} \cline{11-12} \noalign{\smallskip}
		& \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard}  \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
		8  & 11,19 & 6,42  & & 3,98  & 4,01  & & 4,20 & 3,52  & & 10,80 & 10,28 \\
		16 & 5,02  & 7,43  & & 5,07  & 6,76  & & 6,77 & 7,77  & & 15,93 & 18,46 \\
		32 & 13,59 & 16,45 & & 14,97 & 22,41 & & 7,27 & 15,64 & & 31,96 & 38,76 \\
		64 & 30,42 & 7,70  & & 14,49 & 23,48 & & 5,79 & 22,61 & & 31,65 & 30,57 \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

>>>>

%Поскольку в реальных задачах глобальной оптимизации каждое вычисление значения функции (далее \textit{trial}) является операцией, требующей больших вычислительных затрат, количество таких попыток приходится сокращать. Это может быть достигнуто за счет преднамеренного выбора вариантов в ходе поиска оптимального решения, отсекая бесперспективные поисковые под области и исследуя только те, в которых можно найти решение задачи. Алгоритм глобального поиска (АГП) основан на этой идее \cite{Strongin2000}. В настоящей работе мы попытались объединить АГП и метод локальной оптимизации \cite{Nocedal, Kelley}  (метод поиска по образцу Хука-Дживса \cite{HookJeeves}), чтобы уменьшить количество выполняемых испытаний. Решение о запуске локального метода будет приниматься с использованием дерева решений.
In real-world global optimization problems, each function evaluation (hereafter referred to as a \textit{trial}) is a computationally expensive operation, making it necessary to reduce the number of such trials. This may be achieved by the purposeful selection of alternatives during the course of optimization, rejecting unpromising search sub-regions and exploring only those where a solution can be found. The Global Search Algorithm (GSA) is based on this idea \cite{Strongin2000}. In this work, we attempted to combine the GSA and a local optimization method (the Hooke-Jeeves pattern search method \cite{Nocedal, Kelley}) to reduce the number of trials performed. The decision to run the local method will be made using a decision tree.

@@

Classical local optimization algorithms are designed to find a single extremum (local minimum or maximum) in the neighborhood of an initial guess. Depending on the information used about the objective function, these methods are typically classified as:
\begin{itemize}
	\item Zero-order methods, which use only function values;
	\item First-order methods (gradient methods), requiring the computation of first derivatives;
	\item Second-order methods (e.g., Newton's method), which require the computation of the matrix of second derivatives (Hessian).
\end{itemize}

@@

Algorithms terminated when current trial point was within the $\varepsilon$-vicinity of the true global minimizer. Tables \ref{tab:1} and \ref{tab:2} show the average time and speedup relative to a sequential run, achieved by the algorithms when solving optimization problems. Figure \ref{fig:fig4} shows the speedup in terms of the number of iterations. Parallelization was implemented using MPI, and experiments were conducted on $P = 1, 8, 16, 32, 64$ processes.

@@

\begin{table}
	\caption{Average time to solve the problem}
	\label{tab:1}
	\center
	\begin{tabular}{cccccccccccc}
		\hline\noalign{\smallskip}
		$P$ & \multicolumn{2}{c}{ $N=2$ } & & \multicolumn{2}{c}{$N=3$} & & \multicolumn{2}{c}{$N=4$} & & \multicolumn{2}{c}{$N=5$}  \\
		\noalign{\smallskip} \cline{2-3} \cline{5-6} \cline{8-9} \cline{11-12} \noalign{\smallskip}
		& \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard}  \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
		1   & 6.35 & 12.71 & & 5.92 & 14.19 & & 31.88 & 66.29  & & 53.77 & 324.04 \\
		8   & 0.57 & 1.98  & & 1.63 & 5.59  & & 9.44  & 34.93  & & 6.62 & 52.42 \\
		16  & 1.27 & 1.71  & & 1.17 & 2.10  & & 4.71  & 8.53   & & 3.38 & 17.55 \\
		32  & 0.95 & 0.87  & & 0.38 & 0.75  & & 4.67  & 4.95   & & 1.99 & 11.48 \\
		64  & 0.21 & 1.65  & & 0.41 & 0.60  & & 5.51  & 2.93   & & 1.70 & 10.60 \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}

\begin{table}
	\caption{Time speedup}
	\label{tab:2}
	\center
	\begin{tabular}{cccccccccccc}
		\hline\noalign{\smallskip}
		$P$ & \multicolumn{2}{c}{ $N=2$ } & & \multicolumn{2}{c}{$N=3$} & & \multicolumn{2}{c}{$N=4$} & & \multicolumn{2}{c}{$N=5$}  \\
		\noalign{\smallskip} \cline{2-3} \cline{5-6} \cline{8-9} \cline{11-12} \noalign{\smallskip}
		& \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard} & & \textit{Simple} & \textit{Hard}  \\
		\noalign{\smallskip} \hline \noalign{\smallskip}
		8  & 11.19 & 6.42  & & 3.98  & 4.01  & & 4.20 & 3.52  & & 10.80 & 10.28 \\
		16 & 5.02  & 7.43  & & 5.07  & 6.76  & & 6.77 & 7.77  & & 15.93 & 18.46 \\
		32 & 13.59 & 16.45 & & 14.97 & 22.41 & & 7.27 & 15.64 & & 31.96 & 38.76 \\
		64 & 30.42 & 7.70  & & 14.49 & 23.48 & & 5.79 & 22.61 & & 31.65 & 30.57 \\
		\noalign{\smallskip}\hline
	\end{tabular}
\end{table}
